The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-10 kalman_walk_paper.tex
put computational exploration in a subdir

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}
\newcommand{\grad}{\nabla}

@@

\paragraph{Completing the square}
First note that if $A$ is symmetric,
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so if $B$ is also symmetric and $A+B$ is invertible,
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + y^T A y - y^T A^T (A+B)^{-1} A y .
\end{align*}
Since the second line doesn't depend on $x$,
by substituting $A=\Sigma^{-1}$ and $B=U^{-1}$,
\begin{align*}
    \frac{ f(x;\Sigma,y) f(x;U,0) }{\int f(z;\Sigma,y) f(z;U,0) dz}
    &=
    f(x; (\Sigma^{-1} + U^{-1})^{-1}, (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} y) .
\end{align*}

@@

\plr{Fix this calculation to include derivatives wrt to $B$ and $C$ also.}

@@

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $A_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A) 
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_{A_0}(t) \right|^2 dt \\
        &:= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C e^{A_0 t} B \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
How does this change with $A$?
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ)\vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt \\
      &=
        2 \int_0^\infty \rho(t) C \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B \left( h_A(t) - h_{A_0}(t) \right)^T dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_{A_0}(t)$, (so wolog $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH(Y,Z) &:= \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ)\vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) C 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B B^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C^T dt  .
  \end{aligned}
\end{equation}
Here $\calH$ is the quadratic form underlying the Hamiltonian.
By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
and 0 elsewhere,
the coefficients of the quadratic form is
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &:=
        \calH(\Delta_{ij}, \Delta_{k\ell}) .
    \end{aligned}
\end{equation}

>>>>

\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\var}{\mathop{\mbox{Var}}}
\newcommand{\cov}{\mathop{\mbox{Cov}}}
\newcommand{\tr}{\mathop{\mbox{tr}}} % trace
\newcommand{\pda}{\frac{\partial}{\partial A_{ij}}}
\newcommand{\ind}{\mathds{1}}
\newcommand{\grad}{\nabla}

@@

\paragraph{Completing the square}
First note that if $A$ is symmetric,
\begin{align*}
    (x-y)^T A (x-y)
    &=
    x^T A \left( x - 2y \right) + y^T A y ,
\end{align*}
and so if $B$ is also symmetric and $A+B$ is invertible,
\begin{align*}
    (x-y)^T A (x-y) + x^T B x
    &=
    x^T (A + B) \left( x - 2 (A + B)^{-1} A y \right) + y^T A y \\
    &=
    \left( x - (A + B)^{-1} A y \right)^T
    (A + B)
    \left( x - (A + B)^{-1} A y \right) \\
    &\qquad{}
    + y^T A y - y^T A^T (A+B)^{-1} A y .
\end{align*}
Therefore, 
by substituting $A=\Sigma^{-1}$ and $B=L^T L$,
\begin{align*}
    \frac{ f(x;\Sigma,y) \exp(-x^T L^T L x / 2)}{\int f(z;\Sigma,y) \exp(-z^T L^T L z / 2) dz}
    &=
    f(x; (\Sigma^{-1} + L^T L)^{-1}, (\Sigma^{-1}+L^T L)^{-1} \Sigma^{-1} y) .
\end{align*}

@@

@@

Suppose that $\rho(t) \ge 0$ is a weighting function on $[0,\infty)$
so that fitness is a function of $L^2(\rho)$ distance of the impulse response from optimal.
With $h_0(t) = C_0 e^{tA_0} B_0$ a representative of the optimal set:
\begin{equation}
    \begin{aligned}
        D(A, B, C) 
        &:= 
        \int_0^\infty \rho(t) \left| h_A(t) - h_0(t) \right|^2 dt \\
        &:= 
        \int_0^\infty \rho(t) \left| C e^{At} B - C_0 e^{A_0 t} B_0 \right|^2 dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
        \right\} dt \\
        &= 
        \int_0^\infty \rho(t) \tr\left\{
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)
            \left( C e^{At} B - C_0 e^{A_0 t} B_0 \right)^T
        \right\} dt  .
        % &= 
        % \int_0^\infty \rho(t)  \tr\left\{
        %         C e^{At} B B^T e^{A^T t} C^T 
        %         - C_0 e^{A_0t} B_0 B^T e^{A^T t} C^T 
        %         \right. \\ &\qquad \qquad \left. {}
        %         - C e^{At} B B_0^T e^{A_0^T t} C_0^T 
        %         + C_0 e^{A_0t} B_0 B_0^T e^{A_0^T t} C_0^T 
        % \right\} dt \\
        % \int_0^\infty \rho(t) \left| C \left( e^{At} - e^{A_0 t} \right) B \right|^2 dt \\
        % &= 
        % \int_0^\infty \rho(t) C \left( e^{At} - e^{A_0 t} \right) B B^T \left( e^{At} - e^{A_0 t} \right)^T C^T dt
    \end{aligned}
\end{equation}
How does this change as we perturb about $(A_0, B_0, C_0)$?
First we differentiate with respect to $A$, keeping $B=B_0$ and $C=C_0$ fixed.
Since
\begin{equation}
  \begin{aligned}
      \frac{d}{du} e^{(A+uZ)t} \vert_{u=0}
      &=
      \int_0^t e^{As} Z e^{A(t-s)} ds, 
  \end{aligned}
\end{equation}
we have that
\begin{equation}
  \begin{aligned}
      \frac{d}{du} D(A+uZ,B_0,C_0)\vert_{u=0}
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 B_0^T \left( e^{At} - e^{A_0 t} \right)^T C_0^T \right\} dt \\
      &=
        2 \int_0^\infty \rho(t) \tr\left\{ C_0 \left( \int_0^t e^{As} Z e^{A(t-s)} ds \right) B_0 \left( h_A(t) - h_0(t) \right)^T \right\} dt 
  \end{aligned}
\end{equation}
and, by differentiating this and supposing that $A$ is on the optimal set,
i.e., $h_A(t)=h_0(t)$, (so wolog $A=A_0$):
\begin{equation}
  \begin{aligned}
      \calH^{A,A}(Y,Z) 
      &:= 
      \frac{1}{2} \frac{d}{du} \frac{d}{dv} D(A_0+uY+vZ,B_0,C_0)\vert_{u=v=0} \\
      &=
        \int_0^\infty \rho(t) \tr\left\{ C_0 
        \left( \int_0^t e^{A_0 s} Y e^{A_0 (t-s)} ds \right) 
        B_0 B_0^T 
        \left( \int_0^t e^{A_0 s} Z e^{A_0 (t-s)} ds \right)^T
        C_0^T \right\} dt  .
  \end{aligned}
\end{equation}

The function $\calH$ will define a quadratic form.
To illustrate the use of this, suppose that $B$ and $C$ are fixed.
% Here $\calH$ is the quadratic form underlying the Hamiltonian.
By defining $\Delta_{ij}$ to be the matrix with a 1 in the $(i,j)$th slot
and 0 elsewhere,
the coefficients of the quadratic form is
\begin{equation}
    \begin{aligned}
        H_{ij, k\ell}(A)
        &:=
        \calH(\Delta_{ij}, \Delta_{k\ell}) .
    \end{aligned}
\end{equation}
